---
title: "BITHub datasets"
author: "Urwah Nawaz"
date: "2023-06-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background 

This documentation contains the summary of datasets that are currently hosted on BITHub. It also includes levels of information (gene, transcript, exon) and overall summary statistics. 

```{r}
source("functions.R")
library(dplyr)
library(purrr)
library(recount3)
```




# Bulk datasets on BITHub 

* PE 
* GTEx 
* BrainSpan
* BrainSeq 

```{r}
bspan.md = read.csv("/home/neuro/Documents/BrainData/Bulk/BrainSpan/Formatted/BrainSpan-metadata-subset.csv", 
                    header=TRUE)
bseq.md = read.csv("/home/neuro/Documents/BrainData/Bulk/Brainseq/Formatted/BrainSeq-metadata-subset.csv", 
                   header=TRUE)
pe.md = read.csv("/home/neuro/Documents/BrainData/Bulk/PsychEncode/Formatted/PsychEncode-metadata-subset.csv", 
                 header=TRUE)
gtex.md = read.csv("/home/neuro/Documents/BrainData/Bulk/GTEx/Formatted/GTEx-metadata-subset.csv", header=TRUE)
```

```{r}
# Table showing total number of 
# Samples
# Age Intervals 
# Regions 
# Sample normalization

```

```{r}
genres <- c("Action", "Adventure", "Science Fiction", "Thriller", "Western", "Drama", "Crime", "Comedy")

movie_data <- tibble(
  genres = map_chr(sample(5, 50, replace = TRUE), function(x){
    sample(genres, size = x, replace = FALSE) %>%
      paste(collapse = "|")
  }),
  vote_average = runif(length(genres), 0, 10)
)

summary

bulk_datasets = list(bspan.md, bseq.md, gtex.md, pe.md)

summary = data.frame(Dataset = c("BrainSpan"), 
                     Samples = c(nrow(bspan.md)), 
                     "Age Interval" = c(unique(bspan.md$AgeInterval) %>% 
                                          paste(collapse = ",") ), 
                     Regions = c(unique(bspan.md$StructureAcronym) %>% paste(collapse = ",")), 
                     PMI = c("TRUE"), 
                     RIN = c("TRUE"), 
                     AgeNumeric = c("TRUE"))

unique(bspan.md$StructureAcronym) %>% paste(collapse = ",")


```

## Proposal for new datasets 

* Gandal et al 

```{r}

load("/home/neuro/Documents/Brain_integrative_transcriptome/Data/Broad-transcriptomic-dysregulation-across-the-cerebral-cortex-in-ASD/data_provided")
```

```{r}
dim(datMeta)
head(datSeq)
head(rsem_gene) ## gene counts 
```
* transcript level data for BrainSeq

```{r}
load("/home/neuro/Documents/Brain_integrative_transcriptome/Data/rse_tx_unfiltered.Rdata")
```

```{r}
head(rse_tx)
## BrainSeq contains transcript counts 
```


## Recount datasets 

* GTEx 
```{r}
gtex = recount3::create_rse_manual(
    project = "BRAIN",
    project_home = "data_sources/gtex",
    organism = "human",
    annotation = "gencode_v26",
    type = "gene"
)


rowData(gtex)
gtex@assays@data@listData$raw_counts
  
```

* Human developmental biology resource 
```{r}
hdbr = recount3::create_rse_manual(
    project = "ERP016243",
    project_home = "data_sources/sra",
    organism = "human",
    annotation = "gencode_v26",
    type = "gene"
)
```

*  Ramaker et al (2017)

```{r}
recount3::create_rse_manual(
    project = "SRP073813",
    project_home = "data_sources/sra",
    organism = "human",
    annotation = "gencode_v26",
    type = "gene"
)

```


* Cardoso-Moreira et al (2019)

```{r}
recount3::create_rse_manual(
    project = "ERP109002",
    project_home = "data_sources/sra",
    organism = "human",
    annotation = "gencode_v26",
    type = "gene"
)
```


# Single-nucleus datasets on BITHub

* Velmheshev
* HCA 

## Proposal for new datasets 

* Cerebellum dataset 

