---
title: "BrainSpan pre-processing"
author: "Urwah Nawaz"
date: "2023-08-26"
output: 
  html_document:
    code_folding: hide
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    autodep = TRUE,
    eval = TRUE,
	echo = FALSE,
	warning = FALSE,
	message = FALSE
)
```

# Set-up

```{r include = FALSE}
source("functions.R")
source("../signatures/libraries.R")
source("../signatures/functions.R")
load("../../DeconRNAShiny/sigsBrain.rda")
library(recount3)
library(magrittr)
library(tibble)
library(reshape2)
library(SummarizedExperiment)
library(corrplot)
library(dplyr)
library(ggvenn)
library(pander)
library(gridExtra)
library(variancePartition)
library(DT)
library(EnsDb.Hsapiens.v86)
library(singscore)
library(AnnotationHub)
library(stargazer)
library(ggfortify)
library(glue)
library(cowplot)
library(broom)
library(glmpca)
library(naniar)
```

```{r annotations}
gene2symbol = genes(EnsDb.Hsapiens.v86, return.type="DataFrame") %>%
    as.data.frame()
gene2symbol %<>% 
    dplyr::select("EnsemblID" = "gene_id", gene_name)

```


```{r include=FALSE}
ah <- AnnotationHub() %>%
    subset(species == "Homo sapiens") %>%
    subset(rdataclass == "EnsDb") %>%
  subset(genome == "GRCh38")
ensDb <- ah[["AH109606"]]
grTrans <- transcripts(ensDb)
trLengths <- exonsBy(ensDb, "tx") %>%
    width() %>%
    vapply(sum, integer(1))
mcols(grTrans)$length <- trLengths[names(grTrans)]

genesGR = genes(ensDb)
transGR = transcripts(ensDb)

mcols(transGR) = mcols(transGR) %>%
  cbind(
    transcriptLengths(ensDb)[rownames(.), c("nexon", "tx_len")]
  )
```


```{r}
mcols(genesGR) <- mcols(genesGR) %>%
  as.data.frame() %>%
  dplyr::select(
    gene_id, gene_name, gene_biotype, entrezid
  ) %>%
  left_join(
    mcols(transGR) %>%
      as.data.frame() %>%
      mutate(
        tx_support_level = case_when(
          is.na(tx_support_level) ~ 1L, 
          TRUE ~ tx_support_level
        )
      ) %>%
      group_by(gene_id) %>%
      dplyr::summarise(
        n_tx = n(),
        longest_tx = max(tx_len),
        ave_tx_len = mean(tx_len),
        gc_content = sum(tx_len*gc_content) / sum(tx_len)
      ) %>%
      mutate(
        bin_length = cut(
          x = ave_tx_len,
          labels = seq_len(10),
          breaks = quantile(ave_tx_len, probs = seq(0, 1, length.out = 11)),
          include.lowest = TRUE
        ),
        bin_gc = cut(
          x = gc_content,
          labels = seq_len(10),
          breaks = quantile(gc_content, probs = seq(0, 1, length.out = 11)),
          include.lowest = TRUE
        ),
        bin = paste(bin_gc, bin_length, sep = "_")
      ),
    by = "gene_id"
  ) %>%
  set_rownames(.$gene_id) %>%
  as("DataFrame")


trans2Gene <- mcols(transGR) %>%
    as.data.frame() %>%
    dplyr::select(tx_id, gene_id) %>%
    dplyr::filter(!is.na(tx_id), !is.na(gene_id)) %>%
    as_tibble()

```

# Background 

In this document, we will perform the pre-processing of the gene expression matrix from the BrainSpan Development Atlas. 


# Preprocessing 

Following analyses and steps will be performed before the data is ready to be exported to Brain Integrative Transcriptome Hub:

- Metadata pre-processing including: 
    - Adding brain region nomenclature
    - Adding developmental stages 
    - Consistent labeling of column names
    - Adding metadata from different resources 
    
- Quality control checks including: 
    - Regions labels (Experimental method)
    - Determining 3' bias that can arise as a result of degradation
    - GC content of the datatset 
    - PCA
  
- Cellular deconvolution of data-set and performing goodness-of-fit
  - Deconvolve using MultiBrain

- Determining drivers of variation

## Metdata pre-processing 
```{r load-data, include=FALSE}
annot = read.csv("/home/neuro/Documents/Brain_integrative_transcriptome/brain_transcriptome/annotations/BrainSpan-metadata-annot.csv", header=TRUE)

dir = file.path("/home/neuro/Documents/BrainData/Bulk/BrainSpan/Kang/genes_matrix_csv")
columns.metadata = read.csv(file.path(dir, "columns_metadata.csv"), header = TRUE)
exp = read.csv(file.path(dir, "expression_matrix.csv"), header= FALSE, row.names= 1)
rows.metadata = read.csv(file.path(dir, "rows_metadata.csv"))
```

The following BrainSpan metadata columns will be used for BITHub: 

```{r col-names}
colnames(columns.metadata) = annot$BITColumnName[match(colnames(columns.metadata),annot$OriginalMetadataColumnName)]

final_annot =annot %>% 
  dplyr::filter(Include..Yes.No....Interest == "Yes") %>% 
  dplyr::select(-c(Include..Yes.No....Interest))

knitr::kable(final_annot)
```

```{r}
message("Adding additional metadata information")
md = columns.metadata %>% 
      mutate(Stage = add_feature(.$Age, stages), 
             Regions = add_feature(.$StructureAcronym, regions), 
             AgeInterval = add_feature(.$Age, age_intervals), 
             Diagnosis = "Control", 
             Age = gsub(" ","_", .$Age)) %>%  
      mutate(SampleID = paste(DonorID, Age, StructureAcronym, Stage, sep = "_"), 
             age_for_mRIN = gsub("_", "", .$Age), 
             DonorName = gsub("\\.","_", .$DonorName)) %>% 
      mutate('sample name' = paste(DonorName, age_for_mRIN,Sex ,StructureAcronym, sep = "//")) %>% 
      dplyr::select("SampleID", everything())

md$AgeNumeric[grepl("pcw", md$Age, ignore.case = TRUE)]<-
      md$Age[grepl("pcw", md$Age)] %>%  str_remove("_pcw")%>% 
      as.numeric() %>% `-` (40) %>% divide_by(52)
md$AgeNumeric[grepl("mos", md$Age, ignore.case = TRUE)] <- 
      md$Age[grepl("_mos", md$Age)] %>%  str_remove("_mos") %>% 
      as.numeric() %>% divide_by(12)
md$AgeNumeric[grepl("yrs", md$Age, ignore.case = TRUE)] <- 
      md$Age[grepl("_yrs", md$Age)] %>%  str_remove("_yrs") %>% 
      as.numeric 
    
md %<>% mutate(Period = ifelse(.$AgeNumeric >= 0, "Postnatal", "Prenatal"), 
                   colname = paste(DonorName, Age, StructureAcronym, sep = "_"))
```

```{r exp-rename}
colnames(exp) = md$SampleID
rownames(exp) <- rows.metadata$ensembl_gene_id
```


```{r}
md.excel = read_excel("../annotations/BrainSpan-additional.xlsx",sheet =2, col_names = TRUE, skip =1) %>% 
      as.data.frame() %>% mutate_at(.vars = "AllenInstituteID", 
                                    .funs = gsub, pattern = "\\.", replacement = "\\_") %>%
      mutate_at(.vars = "Age", .funs = gsub, pattern = "PCW", replacement = "_pcw") %>% 
      mutate_at(.vars = "Age", .funs = gsub, pattern = "M", replacement = "_mos") %>%
      mutate_at(.vars = "Age", .funs = gsub, pattern = "Y", replacement = "_yrs") %>%
      mutate_at(.vars = "Region/Area", .funs = gsub, pattern = "\\/", replacement = "-") %>% 
      mutate(colname = paste(AllenInstituteID, Age, `Region/Area`, sep = "_")) %>% 
      dplyr::select(-c(Agerange, Age, Description))
```

```{r}
md = md %>% 
      left_join(.,md.excel, by = "colname", keep = TRUE)
```

```{r}
md.excel = read_excel("../annotations/BrainSpan-additional.xlsx", sheet = 1, col_names = TRUE) %>% 
      as.data.frame() %>% 
      mutate_at(.vars="Internal ID", .funs = gsub, pattern = "\\.", replacement = "\\_") %>% 
      dplyr::rename("Braincode" = "External ID")

md %<>% 
  left_join(md.excel, by ="Braincode") %>% 
      dplyr::select(-c("Age.y", "colname.y", "colname.x", "Gender", AllenInstituteID,
                     "Region/Area", "age_for_mRIN", "sample name", "Internal ID")) %>% 
      dplyr::rename("Ethnicity"="Ethn.")  %>% 
  distinct(SampleID, .keep_all = TRUE)


   
md = md[!duplicated(md[,c('column_num')]),]
md %>% 
      dplyr::arrange(column_num)
    
```


### Export metadata and expression data 

```{r}
write.csv(exp, file = "/home/neuro/Documents/BrainData/Bulk/BrainSpan/Formatted/BrainSpan-exp.csv")
write.csv(md, file = "/home/neuro/Documents/BrainData/Bulk/BrainSpan/Formatted/BrainSpan-metadata.csv")
```


## Quality control checks 

- Check sample labeling 


- Determining 3' bias that can arise as a result of degradation

```{r}
#read coverage h (average RPKM × L) reflects the abundance and length of each transcript, but not the sequencing #depth of each sample.


```


- GC content of the dataset 

```{r}


```



## Deconvolution 


```{r}
bspan.decon = run_dtg(exp ,sigsBrain$MB) 

gof_res = write.gof(exp, bspan.decon, 
                    signatureUsed = sigsBrain$MB)


gof_res %>% 
    mutate(col = "BrainSpan") %>%
    as.data.frame() %>% 
    rownames_to_column("SampleID") %>% 
    mutate(SampleID = gsub("X", "", SampleID)) %>%
  mutate(SampleID = gsub("\\.", "_", SampleID)) %>%
    left_join(md, by = "SampleID") %>%
    drop_na(Regions) %>%
    ggplot(aes(Regions, r, fill = Regions)) + geom_violin(width=1, alpha = 0.85) + 
  geom_boxplot(width=0.1, color="white") +
  theme_bw() + scale_fill_manual(values = c("Cortex" = "#E85772", 
                                             "Subcortex" = "#878DC5", 
                                             "Cerebellum" = "#FFD357")) + 
  geom_hline(yintercept = c(0.5,0.7), color = "grey",linetype='dotted') + 
  ylab("Goodness of fit (r)")
```

```{r}
gof_res %>% 
    mutate(col = "BrainSpan") %>%
    as.data.frame() %>% 
    rownames_to_column("SampleID") %>% 
    mutate(SampleID = gsub("X", "", SampleID)) %>%
  mutate(SampleID = gsub("\\.", "_", SampleID)) %>%
    left_join(md, by = "SampleID") %>%
  mutate(Period = factor(Period, levels = c("Prenatal", "Postnatal"))) %>%
    drop_na(Regions) %>%
    ggplot(aes(Period, r, fill = Period)) + geom_violin(width=1) + 
  geom_boxplot(width=0.1, color="white") +
  theme_bw() + scale_fill_manual(values = c("Prenatal" = "#F38BA9", 
                                             "Postnatal" = "#46597D")) + 
  geom_hline(yintercept = c(0.5,0.7), color = "grey",linetype='dotted') + 
  ylab("Goodness of fit (r)")
```


## Selection of metadata variables for BITHub 



## Determining drivers of variation 

```{r}
before = log(exp + 0.01) %>%
    melt() %>%
    dplyr::filter(is.finite(value)) %>% 
    ggplot(aes(x=value, color = as.character(variable))) +
    geom_density() +
    ggtitle("Before filtering") +
    labs(x = "logRPKM", y = "Proportion of Genes") + 
    theme_bw() +
    theme(legend.position = "none")

bspan.exp.filt <- thresh(exp)


after =log(bspan.exp.filt + 0.01) %>%
    melt() %>%
    dplyr::filter(is.finite(value)) %>% 
    ggplot(aes(x=value, color = as.character(variable))) +
    geom_density() +
    ggtitle("After filtering") +
    labs(x = "logRPKM", y = "Proportion of Genes") + 
    theme_bw() +
    theme(legend.position = "none")

grid.arrange(before, after, ncol =2)
```

- Principal component analysis 


```{r pca}
pc = log(bspan.exp.filt + 0.05) %>% 
  t() %>% 
   prcomp(scale = TRUE)

pc$x %>% 
  as.data.frame() %>%
  dplyr::select(PC1, PC2, PC3) %>%
   rownames_to_column("SampleID") %>% 
  left_join(md, by = "SampleID") %>%
 # pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value") %>% 
  ggplot(
    aes(PC1, PC2, colour = as.character(StructureAcronym))
  ) +
  geom_point()
```

```{r}
# mcols(transGR) %>%
#   as.data.frame() %>%
#   dplyr::filter(tx_id %in% rownames(pca$rotation)) %>%
#   as_tibble() %>%
#   mutate(
#     bin_length = cut(
#       x = tx_len,
#       labels = seq_len(10),
#       breaks = quantile(tx_len, probs = seq(0, 1, length.out = 11)),
#       include.lowest = TRUE
#     ),
#     bin_gc = cut(
#       x = gc_content,
#       labels = seq_len(10),
#       breaks = quantile(gc_content, probs = seq(0, 1, length.out = 11)),
#       include.lowest = TRUE
#     ),
#     bin = paste(bin_gc, bin_length, sep = "_")
#   ) %>%
#   dplyr::select(tx_id, contains("bin")) %>%
#   mutate(
#     PC1 = pca$rotation[tx_id, "PC1"],
#     PC2 = pca$rotation[tx_id, "PC2"]
#   ) %>%
#   pivot_longer(
#     cols = c("PC1", "PC2"),
#     names_to = "PC",
#     values_to = "value"
#   ) %>%
#   group_by(PC, bin_gc, bin_length, bin) %>%
#   summarise(
#     Size = n(),
#     mean = mean(value),
#     sd = sd(value),
#     t = t.test(value)$statistic,
#     p = t.test(value)$p.value,
#     adjP = p.adjust(p, method = "bonf")
#   ) %>%
#   ggplot(
#     aes(bin_length, bin_gc, colour = t, alpha = -log10(adjP), size = Size)
#   ) +
#   geom_point() +
#   facet_wrap(~PC) +
#   scale_colour_gradient2() +
#   scale_size_continuous(range = c(1, 10)) +
#   labs(
#     x = "Transcript Length",
#     y = "GC Content",
#     alpha = expression(paste(-log[10], p[adj]))) +
#   theme(
#     panel.grid = element_blank(),
#     legend.position = "bottom"
#     ) 
```

- Variance partition
