---
title: "GTEx-preprocessing"
author: "Urwah Nawaz"
date: "2023-08-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set-up

```{r include = FALSE}
source("functions.R")
source("../signatures/libraries.R")
source("../signatures/functions.R")
load("../../DeconRNAShiny/sigsBrain.rda")
library(recount3)
library(magrittr)
library(tibble)
library(reshape2)
library(SummarizedExperiment)
library(corrplot)
library(dplyr)
library(ggvenn)
library(pander)
library(gridExtra)
library(variancePartition)
library(DT)
library(EnsDb.Hsapiens.v86)
library(singscore)
library(AnnotationHub)
library(stargazer)
library(ggfortify)
library(glue)
library(cowplot)
library(broom)
library(glmpca)
library(naniar)
```

```{r}
thresh <- function(x) {
      y <- x > 1
      keep <- rowSums(y) > (ncol(y) / 10)
      return(x[keep,])
    }
```

# Background 

# Preprocessing 

Following analyses and steps will be performed before the data is ready to be exported to Brain Integrative Transcriptome Hub:

- Metadata pre-processing including: 
    - Adding brain region nomenclature
    - Adding developmental stages 
    - Consistent labeling of column names
    - Adding metadata from different resources 
    
- Quality control checks including: 
    - Regions labels (Experimental method)
    - Determining 3' bias that can arise as a result of degradation
    - GC content of the datatset 
    - PCA
  
- Cellular deconvolution of data-set and performing goodness-of-fit
  - Deconvolve using MultiBrain

- Determining drivers of variation

## Metdata pre-processing 

```{r}
annot = read.csv("/home/neuro/Documents/Brain_integrative_transcriptome/brain_transcriptome/annotations/GTEx-metadata-annot.csv", header=TRUE)

dir = file.path("/home/neuro/Documents/BrainData/Bulk/GTEx")

```

```{r}
attributes = list.files(dir, full.names = TRUE, pattern = "\\SampleAttributesDS.txt") # Sample attributes contains sample level information
phenotype = list.files(dir, full.names = TRUE, pattern = "\\SubjectPhenotypesDS.txt") # Phenotype level information related to each donor 
exp = list.files(dir, full.names = TRUE, pattern = "\\gene_tpm.gct.gz") # File used for expression matrix 
    
md = read_tsv(attributes, col_types = c('.default' = 'c')) %>% 
      dplyr::filter(SMTS == 'Brain') %>% 
      mutate(SUBJID = sapply(str_split(SAMPID, pattern = "-"), function(x) paste(x[1:2], collapse = '-'))) %>%
      left_join(read_tsv(phenotype, col_types = c('.default' = 'c')))  %>% as.data.frame()
colnames(md) = annot$BITColumnName[match(colnames(md), annot$OriginalMetadataColumnName)]
    
md = md %>% mutate(StructureAcronym = add_feature(.$Structure, structure_acronym)) %>% 
      mutate(Regions = add_feature(.$StructureAcronym, regions), 
             AgeInterval = paste(.$Age, "yrs", sep = ""), 
             Diagnosis = "Control", 
             Sex = ifelse(Sex == 1, "M", "F"), 
             Period = "Postnatal") %>% as.data.frame()
    
exp = read.delim(exp, skip = 2)
colnames(exp) <- gsub("\\.", "-", colnames(exp)) # Changing expression file names to match metadata SampleIDs
exp %<>% column_to_rownames("Name")
    
exp = exp %>% 
      dplyr::select(contains(md$SampleID))
    message(paste0("Samples subsetted - Exp matrix contains ", ncol(exp), " samples"))
    rownames(exp) <- sub("\\.[0-9]*$", "", rownames(exp))
    
md= md[which(md$SampleID %in% colnames(exp)),] 
```

## Export data
```{r}
write.csv(exp, file = file.path(dir, "Formatted/GTEx-exp.csv"))
write.csv(md, file = file.path(dir,"Formatted/GTEx-metadata.csv"))

```


# Quality control checks 

- Check sample labeling 


- Determining 3' bias that can arise as a result of degradation

```{r}
#read coverage h (average RPKM × L) reflects the abundance and length of each transcript, but not the sequencing #depth of each sample.


```


- GC content of the dataset 

```{r}


```

### Metadata 

```{r}
annot_seq = annot %>% 
  dplyr::filter(Type == "Sequencing metrics")
md.clean = md %>% 
  dplyr::select(SampleID, contains(annot_seq$BITColumnName))
md.clean = md.clean %>% select_if(~ !any(is.na(.)))
md.clean = md.clean[vapply(md.clean, function(x) length(unique(x)) > 1, logical(1L))]
```

- Correlation of metadata variables 

```{r}
M = cor(data.matrix(md.clean), use = "complete.obs")
corrplot(M, order='AOE',
         method = "circle", type = "upper", 
         number.cex = .50, 
         tl.cex=0.75)


```

## Variance partition


```{r}
gtex.vp = read.csv("/home/neuro/Documents/BrainData/Bulk/GTEx/Formatted/GTEx-varPart.csv", 
                   header= TRUE, row.names =1)
gtex_plot = plotVarPart(gtex.vp %>% 
              as.data.frame() %>% 
              dplyr::select(Regions, rRNA_rate,DateofBatch, TypeofBatch, RIN,
                          BSS_Collection_side_code, everything()))



ggsave(gtex_plot, 
       filename = "/home/neuro/Documents/Brain_integrative_transcriptome/Results/Thesis_plots/varPart_gtex.svg",
       height = 6, 
       width = 8, 
       units = "in")
```

#  Deconvolution

```{r}
gtex.decon = run_dtg(exp ,sigsBrain$MB) 
```

```{r}
write.csv(gtex.decon, file = "/home/neuro/Documents/BrainData/Bulk/GTEx/Formatted/GTEx-decon-MB.csv")
```
