---
title: "PsychEncode pre-processing"
author: "Urwah Nawaz"
date: "2023-08-28"
output: 
  html_document:
    code_folding: hide
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    autodep = TRUE,
    eval = TRUE,
	echo = FALSE,
	warning = FALSE,
	message = FALSE
)
```


# Set-up

```{r include = FALSE}
source("functions.R")
source("../signatures/libraries.R")
source("../signatures/functions.R")
load("../../DeconRNAShiny/sigsBrain.rda")
library(recount3)
library(magrittr)
library(tibble)
library(reshape2)
library(SummarizedExperiment)
library(corrplot)
library(dplyr)
library(ggvenn)
library(pander)
library(gridExtra)
library(variancePartition)
library(DT)
library(EnsDb.Hsapiens.v86)
library(singscore)
library(AnnotationHub)
library(stargazer)
library(ggfortify)
library(glue)
library(cowplot)
library(broom)
library(glmpca)
library(naniar)
```


# Background 

# Preprocessing 

Following analyses and steps will be performed before the data is ready to be exported to Brain Integrative Transcriptome Hub:

- Metadata pre-processing including: 
    - Adding brain region nomenclature
    - Adding developmental stages 
    - Consistent labeling of column names
    - Adding metadata from different resources 
    
- Quality control checks including: 
    - Regions labels (Experimental method)
    - Determining 3' bias that can arise as a result of degradation
    - GC content of the datatset 
    - PCA
  
- Cellular deconvolution of data-set and performing goodness-of-fit
  - Deconvolve using MultiBrain

- Determining drivers of variation 

```{r load-data}
annot = read.csv("/home/neuro/Documents/Brain_integrative_transcriptome/brain_transcriptome/annotations/PsychEncode-metadata-annot.csv", header=TRUE)

dir = file.path("/home/neuro/Documents/BrainData/Bulk/PsychEncode/")
exp = list.files(dir, full.names = TRUE, pattern = "\\Gene_expression_matrix_TPM.txt") %>% 
      read.table(., header=TRUE, row.names = 1, check.names = FALSE)
md = list.files(dir, full.names = TRUE, pattern = "Job*") %>% read.csv(., header=TRUE) %>% 
  dplyr::filter(individualID == "2015-1")
comp = list.files(dir, full.names = TRUE, pattern = "\\Cell_fractions*") %>% read_excel() %>%
      as.data.frame() %>% 
      column_to_rownames("CellType")
```

## Metadata
```{r}
colnames(md) = annot$BITColumnName[match(colnames(md), annot$OriginalColumnName)]
    # Fix existing columns 
comp = comp[,-1] 
comp = as.data.frame(t(comp))
m <- match(md$SampleID, rownames(comp))
md <- cbind(md, comp[m,])


    
# PCW to age numeric 
md$AgeNumeric[grepl("PCW", md$AgeNumeric, ignore.case = TRUE)] = md$AgeNumeric[grepl("PCW", md$AgeNumeric)] %>%
  str_remove("PCW")%>% 
  as.numeric() %>% `-` (40) %>% divide_by(52)
    
md$AgeNumeric = gsub("90+", "91", md$AgeNumeric)
md$AgeNumeric =gsub("\\+", "", md$AgeNumeric)
    
    
md$AgeNumeric <- as.numeric(as.character(md$AgeNumeric))
    
md %<>%
      dplyr::filter(Diagnosis == "Affective Disorder" |
               Diagnosis == "Autism Spectrum Disorder" | 
               Diagnosis == "Bipolar Disorder" |
               Diagnosis == "Control" |
               Diagnosis == "Schizophrenia") %>% 
      mutate(Structure = c("Dorsolateral Prefrontal Cortex"),  ## Adding name of structure
             StructureAcronym = c("DLPFC")) %>%  
      mutate(Period = ifelse(.$AgeNumeric >= 0, "Postnatal", "Prenatal"))  %>%
      mutate(Age_rounded = as.character(sapply(.$AgeNumeric, num_to_round))) %>% as.data.frame() %>%
      mutate(AgeInterval = as.character(add_feature(.$Age_rounded, age_intervals))) %>% 
      mutate(Regions = c("Cortex")) %>% 
      mutate(DonorID = as.character(.$SampleID)) %>%
      dplyr::select(-Age_rounded) %>%
      as.data.frame()
    
    
exp %<>% rownames_to_column("EnsemblID")
  
```

## Export
```{r}
write.csv(exp, file = file.path(dir, "Formatted/PsychEncode-exp.csv"))
write.csv(md, file = file.path(dir,"Formatted/PsychEncode-metadata.csv"))

```


